---
title: "580Project"
author: "Zian Shang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r}
# include packages

```


```{r}
# read train.data
train.data <- read.csv("/Users/zianshang/Downloads/AMS\ 580/group\ project\ 580/train.csv", 
                       header = T, 
                       na.strings = " ?")

# remove NA
train.data <- na.omit(train.data)

# remove empty space in front of each char cell
# factorize char columns

# ***if needed, encoding char cells instead of factorizing them***

train.data[] <- lapply(train.data, 
                       function(x) {
                         if(is.character(x)) 
                           as.factor(trimws(x))
                         else x
                       })

View(train.data)
str(train.data)
```


### variable selection, identify important features

correlation analysis-> 1 person

feature importance scores -> 1 person

dimensionality reduction techniques -> 1 person


###
tree-based models -> use education(categorical, factorized)
```{r}
train.data.1 <- subset(train.data, select = -c(education.num))
```

###
logistic, linear, NN models -> use education.num(numeric, original)
```{r}
train.data.2 <- subset(train.data, select = -c(education))
```

*** remember to optimize model parameters***

### model 1: linear model -> 1 person
data scaling

### model 2: multiple linear model -> 1 person
data scaling

### model 3: logistic model -> 1 person
data scaling

### model 4: NN model -> 1 person
data scaling

### model 5: CART model -> 1 person
do not scale

### model 6: random forest model -> 1 person
do not scale

### model 7: extra models, PCA, KNN... -> decide latter





```{r}
# check for data imbalance
table(df$income)
prop.table(table(df$income))
# somewhat imbalanced
```

### ***if model predicts income>50K poorly, try handle imbalance***












```{r}
# read test data
test.data <-read.csv("/Users/zianshang/Downloads/AMS\ 580/group\ project\ 580/test.csv", 
                     header = T, 
                     na.strings = " ?")

# remove NA
test.data <- na.omit(test.data)

# remove empty space in front of each char cell
# ***if needed, factorize/encode char cells***
test.data[] <- lapply(test.data, 
                       function(x) {
                         if(is.character(x)) trimws(x) else x
                       })



View(test.data)
```





#### this line is used to test push to remove self branch & merge data to remove main branch


























