---
title: "580Project"
author: "Zian Shang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r}
# include packages

```


```{r}
# read train.data
train.data <- read.csv("train.csv", 
                       header = T, 
                       na.strings = " ?")

# remove NA
train.data <- na.omit(train.data)

# remove empty space in front of each char cell
# factorize char columns

# ***if needed, encoding char cells instead of factorizing them***

train.data[] <- lapply(train.data, 
                       function(x) {
                         if(is.character(x)) 
                           as.factor(trimws(x))
                         else x
                       })

# View(train.data)
str(train.data)
```


### variable selection, identify important features

correlation analysis-> Tasfia

feature importance scores -> random forest

dimensionality reduction techniques -> Zian
Multicollinearity, PCA Variance

outliers -> Matthew


###
tree-based models -> use education(categorical, factorized)
```{r}
train.data.1 <- subset(train.data, select = -c(education.num))
```

###
logistic, linear, NN models -> use education.num(numeric, original)
```{r}
train.data.2 <- subset(train.data, select = -c(education))
```

*** remember to optimize model parameters***

### model 1: multiple linear model -> Thomas
data scaling

### model 2: logistic model -> Zian
data scaling

### model 3: NN model -> Tasfia
data scaling

### model 4: CART model -> Nicole
do not scale

### model 5: random forest model -> Matthew
do not scale

### model 6: extra models, PCA, KNN... -> decide latter

### explore XgBoost classification --> Thomas





```{r}
# check for data imbalance
table(df$income)
prop.table(table(df$income))
# somewhat imbalanced
```

### ***if model predicts income>50K poorly, try handle imbalance***












```{r}
# read test data
test.data <-read.csv("test.csv", 
                     header = T, 
                     na.strings = " ?")

# remove NA
test.data <- na.omit(test.data)

# remove empty space in front of each char cell
# ***if needed, factorize/encode char cells***
test.data[] <- lapply(test.data, 
                       function(x) {
                         if(is.character(x)) trimws(x) else x
                       })


View(test.data) # tasfia's comment check
```





#### this line is used to test push to remove self branch & merge data to remove main branch ####



nicki testing 2






















