---
title: "580Project"
author: "Zian Shang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r}
# include packages
library(tidyverse)
library(caret)
library(corrplot)
library(randomForest)
library(nnet)
library(rpart)
library(rpart.plot)
library(xgboost)
library(neuralnet)
```

```{r}
# read train.data
train.data <- read.csv("train.csv", 
                       header = T, 
                       na.strings = " ?") 

# check for NA
colSums(is.na(train.data))

###### should we impute instead of remove NA?

# remove NA
train.data <- na.omit(train.data)

# remove empty space in front of each char cell
# factorize char columns

# ***if needed, encoding char cells instead of factorizing them***

train.data[] <- lapply(train.data, 
                       function(x) {
                         if(is.character(x)) 
                           as.factor(trimws(x))
                         else x
                       })
# view(train.data)
str(train.data)
```
```{r}
# read test data
test.data <-read.csv("test.csv", 
                     header = T, 
                     na.strings = " ?")

# remove NA
test.data <- na.omit(test.data)

# remove empty space in front of each char cell
# ***if needed, factorize/encode char cells***

test.data[] <- lapply(test.data, 
                       function(x) {
                         if(is.character(x)) 
                           as.factor(trimws(x))
                         else x
                       })


# View(test.data) 
str(test.data)
```
###
tree-based models -> use education(categorical, factorized)
```{r}
train.data.1 <- subset(train.data, select = -c(education.num))
test.data.1 <- subset(train.data, select = -c(education.num))
```

###
logistic, linear, NN models -> use education.num(numeric, original)
```{r}
train.data.2 <- subset(train.data, select = -c(education))
test.data.2 <- subset(train.data, select = -c(education))
```

### variable selection, identify important features

correlation analysis -> Tasfia

Correlation Matrix for Numerical Variables:
``` {r}
# Identify numeric columns
numeric_vars <- names(train.data)[sapply(train.data, is.numeric)]

# Subset only numeric columns
numeric_data <- train.data[, numeric_vars]

# Compute correlation matrix
corr_matrix <- cor(numeric_data, method = "pearson")
corr_matrix
```

``` {r}
# Display heat map
corrplot(cor_matrix,
         method = "color",
         col = colorRampPalette(c("white", "lightpink", "deeppink4"))(10),
         tl.col = "black",
         tl.cex = 0.9,
         addCoef.col = "black",
         number.cex = 0.7)
```
Correlation Matrix for Categorial Variables (just testing this out):
``` {r}
library(vcdExtra)

# Select categorical columns
cat_vars <- train.data[, sapply(train.data, is.factor)]

# Function to compute Cramér's V matrix
cramers_v_matrix <- function(df) {
  vars <- names(df)
  mat <- matrix(NA, ncol = length(vars), nrow = length(vars), 
                dimnames = list(vars, vars))
  for (i in vars) {
    for (j in vars) {
      tbl <- table(df[[i]], df[[j]])
      mat[i, j] <- assocstats(tbl)$cramer
    }
  }
  return(as.data.frame(mat))
}

cramers_v_df <- cramers_v_matrix(cat_vars)
cramer_mat <- as.matrix(cramers_v_df)
corrplot(cramer_mat,
         method = "color",
         col = colorRampPalette(c("white", "lightblue", "darkblue"))(200),
         type = "full",
         tl.col = "black",
         tl.cex = 0.7,
         na.label = " ",
         addCoef.col = "black",
         number.cex = 0.6)
```

Fitting Random Forest Model before feature selection:
``` {r}
set.seed(123)
rf_model <- randomForest(income ~ ., data = train.data.1, importance = TRUE, ntree = 500)

# Predictions on training data
train_preds <- predict(rf_model, newdata = test.data.1)

# Confusion matrix
cm <- confusionMatrix(train_preds, test.data.1$income, positive = ">50K")
cm
```

Feature importance scores using random forest:
``` {r}
# Extract importance scores
importance_df <- as.data.frame(importance(rf_model))

# Add variable names as columns
importance_df$Variable <- rownames(importance_df)

# Reorder columns
importance_df <- importance_df[, c("Variable", "MeanDecreaseAccuracy", "MeanDecreaseGini")]

# Sort by MeanDecreaseGini (most commonly used)
importance_df <- importance_df[order(-importance_df$MeanDecreaseGini), ]
importance_df

# Plot variable importance
varImpPlot(rf_model, 
           main = "Random Forest Variable Importance (All Features)")
```

dimensionality reduction techniques -> Zian

outliers -> Matthew



*** remember to optimize model parameters***

### model 1: multiple linear model -> Thomas
data scaling

### model 2: logistic model -> Zian
data scaling

### model 3: NN model -> Tasfia
data scaling

### model 4: CART model -> Nicole
do not scale

### model 5: random forest model -> Tasfia
do not scale

Random Forest with Reduced Features:
``` {r}
train.rf.reduced <- subset(train.data.1, select = -c(race, sex, native.country, capital.loss))

set.seed(123)
rf_reduced <- randomForest(income ~ ., 
                           data = train.rf.reduced, 
                           ntree = 500, 
                           importance = TRUE)

importance(rf_reduced)

# Plot variable importance
varImpPlot(rf_reduced, 
           main = "Random Forest Variable Importance (Reduced Features)")
```
Confusion Matrix for RF w/ reduced features:
``` {r}
test.rf.reduced <- subset(test.data.1, select = -c(race, sex, native.country, capital.loss))

preds_reduced <- predict(rf_reduced, newdata = test.rf.reduced)

cm <- confusionMatrix(preds_reduced, test.data.1$income, positive = ">50K")
cm
```

### model 6: extra models, PCA, KNN... -> decide latter

### explore XgBoost classification --> Thomas


```{r}
# check for data imbalance
table(df$income)
prop.table(table(df$income))
# somewhat imbalanced
```

### ***if model predicts income>50K poorly, try handle imbalance***











#### this line is used to test push to remove self branch & merge data to remove main branch ####























